---
title: "Economic and Human Cost of Severe Weather Events"
author: "Scott Cogswell"
date: "13 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

This paper explores the effects of severe weather events in the United States.
In particular, we examine the human cost (in fatalities) and the economic cost 
(in monetary damages to property and crops) 
of severe weather using data from the National Oceanic and Atmospheric Administration (NOAA).
We find that tornadoes are the most deadly severe weather phenomenon by a large margin,
and that floods have caused the most damage.

# Data Processing

We use the publicly available Storm Data report file provided by NOAA.
This is a comma separated file which indicates the state or region and county, 
begin and end time, the event type, the numbers of fatalaties and injuries, 
the amount of property damage, the amount of crop damage, and the precise location.


```{r Load Data, cache=TRUE}
url.StormData <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
bz2.StormData <- basename(url.StormData)
if (file.exists(bz2.StormData)) {
  cat("Using file downloaded at", format(file.info(bz2.StormData)$mtime, "%Y-%m-%d %H:%M:%S"))
} else {
  download.file(url.StormData, (bz2.StormData))
}
stormData <- read.csv(bz2.StormData)
stormData[1, ]
stormData[nrow(stormData), ]
```

The file contains 902,297 distinct observations, it may take some time to load.

## Data Coverage

Our storm data file covers a wide variety of events, from 1950 through 2011, 
throughout the United States and territories.
The first observation is a tornado at 01:30 CST on April 18, 1950 followed by 
another tornado at 01:45 CST on the same day in another county in Alabama.
The last observation is also in Alabama, at 08:00:00PM CST on November 28, 2011.

The regions covered are

* All 50 US states
* Washington, DC
* Puerto Rico
* Guam
* American Samoa
* US Virgin Islands

The following oceans and inland waterways are also covered:

* The Great Lakes (each separately)
* The St. Lawrence River
* Pacific Ocean
* Atlantic Ocean
* Gulf of Alaska
* Gulf of Mexico
* Caribbean Sea and Tropical Atlantic

Events are recorded for their possibility of causing 

* loss of life
* injury 
* damage to property or crops or loss of commerce

In our analysis we focus on loss of life and property/crop damage.

Rare and unusual weather and other significant meteorological events 
such as record maximum and minimum temperatures are also recorded.

## Damage scale conventions

Property damage values are recorded in two fields: the first field contains up to
three significant figures, and the second contains the exponent.
PROPDMG and PROPDMGEXP indicate the amount of property damage, while
CROPDMG and CROPDMGEXP indicate the amount of crop damage.
We combine property and crop damage in this analysis.

For example PROPDMG = 25.0 and PROPDMGEXP = M indicates $25 million of property damage.
To assist analysis, we convert the values into numbers by multiplying 
the original values according to the exponent.
The function used is:

```{r DamageScale}
ExpVal <- function(exponent) {
  # Convert an exponent into a value to scale the damage estimate
  # Valid exponents (case insensitive) are H, K, M, and B,
  # other values receive a scale of 1.
  # Damage estimate is this value times the DMG field value.
  
   switch(toupper(exponent),
          H=100, K=1000, M=1000000, B=1000000000, 1)
}
```

A very small proportion of the PROPDMGEXP and CROPDMGEXP values do not fit the 
documented standard of 'K' for thousands, 'M' for millions, and 'B' for billions of dollars.
We assume that 'H' is for hundreds of dollars, although given that the events of
interest are in the tens to hundreds of billions of dollars of damage our findings
are robust to this assumption.

```{r PropertyEXPvalues}
table(stormData$PROPDMGEXP)
# percentage of non-blank observations
table(stormData$PROPDMGEXP[stormData$PROPDMGEXP!=""]) / sum(stormData$PROPDMGEXP!="") * 100
```

Over half of the observations have an empty PROPDMGEXP value, 
over 97% of the remaining value are 'K', 2.6% are 'M' or 'm', 
there are 40 'B', six 'H' and one 'h'. 

The crop data observations follow the published guidelines more closely.

```{r CropEXPvalues}
table(stormData$CROPDMGEXP)
# percentage of non-blank observations 
table(stormData$CROPDMGEXP[stormData$CROPDMGEXP!=""]) / sum(stormData$CROPDMGEXP!="") * 100
```

Fewer than 0.1% of the observations have PROPDMGEXP or CROPDMGEXP which we are unable to interpret,
these are left with a scale of 1.

We calculate the total damage as follows:

```{r ConvertValues}
stormData$Damage <- stormData$PROPDMG * sapply(stormData$PROPDMGEXP, ExpVal) +
                    stormData$CROPDMG * sapply(stormData$CROPDMGEXP, ExpVal)
```

## Injuries and Fatalities

Injuries and fatalaties are recorded for each event, values are zero or a positive number.
NOAA has instructions about recording injuries and fatalities that are caused directly by
the event, and those that are indirectly caused by the event. 
For example, wind blows a tree next to a road onto a passing car killing the driver: this is a direct fatality.
Another car strikes the same fallen tree some time later: 
any injuries or fatalities from this incident are indirect.

## Recording Events

The event type is recorded in the EVTYPE field in the input file,
but this field is not well standardized.
There 985 distinct values in EVTYPE, 
of which 216 distinct EVTYPE values contain "WIND"; these include
"THUNDERSTORM WINDS53", "THUNDERSTORM WINDSS", 
"THUNDERSTORMS WIND", "THUNDERSTORMS WINDS", "THUNDERSTORMW WINDS", "THUNDERSTORMWINDS", "THUNDERSTROM WIND",
"THUNDERSTROM WINDS", "THUNDERTORM WINDS", "THUNDERTSORM WIND", "THUNDESTORM WINDS", and "THUNERSTORM WINDS" among others.
In this analysis we make no effort to combine these beyond what is described below.

# Results

We consider the effects of severe weather events on human health, and in economic consequences
as measured by property and crop damages.

Our first step is to identify the events leading to the greatest effects,
and given the imprecise event names, check that our results are
robust to near matches.
To do this we define a function to show the results of events sharing
a common pattern or regular expression.

```{r Analysis}
evFatal <- sort(tapply(stormData$FATALITIES, stormData$EVTYPE, sum), decr=T)
evFatal <- evFatal[evFatal > 0]   # only the event types with any fatalities
evDamage <- sort(tapply(stormData$Damage, stormData$EVTYPE, sum), decr=T)
evDamage <- evDamage[evDamage > 0]  # only event types with damage

HowMany <- function(ev, ptn, ic=TRUE) {
  vv <- ev[grep(ptn, names(ev), value=T, ignore.case=ic)]; 
  show(vv); 
  unlist(list(Total=sum(vv)))
}
```

With function *HowMany* we can test whether, for example, thunderstorms are
under-represented in our initial list of fatality causes because fatalities are
split across different categories.

## Human health

We measure the effect on human health using fatalities. 
Figure 1 shows the fatality count for each event type.

```{r PlotFatalities}
evFatal <- sort(tapply(stormData$FATALITIES, stormData$EVTYPE, sum), decr=T)
evFatal <- evFatal[evFatal > 0]   # only the event types with any fatalities
old.mar <- par(mar=c(5,9,4,2))
barplot(evFatal[11-(1:10)], horiz=TRUE, mar=c(5,9,4,2), main="Severe Event Fatalities, 1950-2011", xlab="Fatalities", las=2)
par(mar=old.mar)
```

Tornadoes appear to be the most significant cause of death by a large margin,
with 5,633 compared to Excessive Heat with 1,902  
but the ambiguous naming may affect these results as "HEAT" and "WIND" both appear twice in
the top ten.
This also suggests that the 5,633 attributed to TORNADO may also be an understatement.
We examine this possibility by calculating the total fatalities across all similar
event types, using function *HowMany()* as defined above.

```{r FatalEvents}
HowMany(evFatal, "TORN")
HowMany(evFatal, "HEAT")
HowMany(evFatal, "FLASH")
HowMany(evFatal, "FLOOD")
HowMany(evFatal, "LIGHTNING")
HowMany(evFatal, "WIND")
```

The 5,633 deaths attributed to "TORNADO" are not the complete story, but they are very close.
The 1,903 deaths attributed to "EXCESSIVE HEAT", however, are only 60% of the 
deaths attributable to heat related weather: 
other classifications of extreme heat account for another 1,235 fatalities.
Similarly, including fatalities attributed to other forms of flooding increases 
the total from 987 to 1,035 for other classifications of flash floods, 
or 1,525 for all types of flood.

The "TSTM WIND" (assumed to mean thunderstorm wind) type as part of a broader 
"WIND" category also shows a large variety of sub-types, with a total of 
1,451 fatalities, putting the broad wind category ahead of lightning, 
but we note that 25 of these are shared with tornadoes in the 
"TORNADOES, TSTM WIND, HAIL" event type. 
One could also question whether "Whirlwind" should be classified with tornadoes.

## Property and crop damage

To measure the economic impact of severe weather we combine the total damage to
property and crops for each event type as shown in the figure below.

```{r PlotDamage}
evDamage <- sort(tapply(stormData$Damage, stormData$EVTYPE, sum), decr=T) / 1e9
evDamage <- evDamage[evDamage > 0]   # only the event types with any fatalities
old.mar <- par(mar=c(5,11,4,2))
barplot(evDamage[11-(1:10)], horiz=TRUE, mar=c(5,9,4,2), main="Severe Event Damage, 1950-2011", xlab="USD billions", las=2)
par(mar=old.mar)
```

The positions of floods and tornadoes are reversed, with floods taking the top
spot by a significant margin; including other flood types ("FLASH FLOOD" and 
"RIVER FLOOD" are also in the top 10) increases from $150 billion to $179.9 billion
for all types of flooding. This is over three times the $59 billion in damages
attributed to tornadoes, the leading killer. Hurricanes and typhoons replace 
heat in second spot with $90.9 billion in damages, slightly over half the total
flood damage.
Although storm surges are associated with hurricanes, we retain the separate
classification here; combining them would remain second place behind all 
floods.

```{r DamageTotal}
HowMany(evDamage, "FLOOD")
HowMany(evDamage, "(HURR)|(TYP)")
HowMany(evDamage, "TORN")
HowMany(evDamage, "SURGE")
HowMany(evDamage, "HAIL")
HowMany(evDamage, "WIND")
HowMany(evDamage, "ICE")
```

Althogh no individual wind event type makes the top ten, considered together they 
account for $19.7 billion of damage, 
surpassing the $15 billion in drought damage and $9 billion in ice-related damage, 
nearly matching the $20.7 billion from hail.
