---
title: "Economic and Human Cost of Severe Weather Events"
author: "Scott Cogswell"
date: "13 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

This paper explores the effects of severe weather events in the United States.
In particular, we examine the economic (monetary) cast and the human cost (in fatalities and injuries)
of severe weather using data from the National Oceanic and Atmospheric Administration (NOAA).
We find that 


# Data Processing

We use the publicly available Storm Data report file provided by NOAA.
This is a comma separated file which indicates the state or region and county, 
begin and end time, the event type, the numbers of fatalaties and injuries, 
the amount of property damage, the amount of crop damage, and the precise location.


```{r Load Data, cache=TRUE}
url.StormData <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
bz2.StormData <- basename(url.StormData)
if (file.exists(bz2.StormData)) {
  cat("Using file downloaded at", format(file.info(bz2.StormData)$mtime, "%Y-%m-%d %H:%M:%S"))
} else {
  download.file(url.StormData, (bz2.StormData))
}
stormData <- read.csv(bz2.StormData)
```

The file contains 902,297 distinct observations, it may take some time to load.

## Data Coverage

Our storm data file covers a wide variety of events, from 1950 through 2011, 
throughout the United States and territories.
The regions covered are

* All 50 states
* Washington DC
* Puerto Rico
* Guam
* American Samoa
* US Virgin Islands

The following oceans and inland waterways are also covered:

* The Great Lakes
* The St. Lawrence River
* Pacific Ocean
* Atlantic Ocean
* Gulf of Alaska
* Gulf of Mexico
* Caribbean Sea and Tropical Atlantic

Events are recorded for their possibility of causing 

* loss of life
* injury 
* damage to property or crops or loss of commerce

Rare and unusual weather and other significant meteorological events 
such as record maximum and minimum temperatures are also recorded.

## Damage scale conventions

Property damage is recorded in two fields: the first field contains up to
three significant figures, and the second contains the exponent.
PROPDMG and PROPDMGEXP indicate the amount of property damage, while
CROPDMG and CROPDMGEXP indicate the amount of crop damage.
We combine property and crop damage in this analysis.

For example PROPDMG = 25.0 and PROPDMGEXP = M indicates $25 million of property damage.
To assist analysis, we convert the values into numbers by multiplying 
the original values according to the exponent.
The function used is:

```{r DamageScale}
ExpVal <- function(exponent) {
  # Convert an exponent into a value to scale the damage estimate
  # Valid exponents (case insensitive) are H, K, M, and B,
  # other values receive a scale of 1.
  # Damage estimate is this value times the DMG field value.
  
   switch(toupper(exponent),
          H=100, K=1000, M=1000000, B=1000000000, 1)
}
```

A very small proportion of the PROPDMGEXP and CROPDMGEXP values do not fit the 
documented standard:

```{r EXPvalues}
table(stormData$PROPDMGEXP)
table(stormData$PROPDMGEXP[stormData$PROPDMGEXP!=""]) / sum(stormData$PROPDMGEXP!="") * 100

table(stormData$CROPDMGEXP)
table(stormData$CROPDMGEXP[stormData$CROPDMGEXP!=""]) / sum(stormData$CROPDMGEXP!="") * 100
```

Over half of the observations have an empty PROPDMGEXP value, 
over 97% of the remaining value are 'K', 2.6% are 'M' or 'm', 
there are 40 'B', six 'H' and one 'h'. 
Fewer than 0.1% of the observations have PROPDMGEXP or CROPDMGEXP which we are unable to interpret,
these are left with a scale of 1.

```{r ConvertValues}
stormData$Damage <- stormData$PROPDMG * sapply(stormData$PROPDMGEXP, ExpVal) +
                    stormData$CROPDMG * sapply(stormData$CROPDMGEXP, ExpVal)
```

## Injuries and Fataluties

Injuries and fatalaties are recorded for each event, values are zero or a positive number.
NOAA has instructions about recording injuries and fatalities that are caused directly by
the event, and those that are indirectly caused by the event. 
For example, wind blows a tree onto a road onto a passing car killing the driver: this is a direct fatality.
Another car strikes thee same fallen tree some time later: any injuries or fatalities are indirect.

## Recording Events

The event type (EVTYPE) field in the input file is not well standardized.
For example, there 216 distinct EVTYPE values containing "WIND", including
"THUNDERSTORM WINDS53", "THUNDERSTORM WINDSS", "THUNDERSTORMS WIND", "THUNDERSTORMS WINDS",
"THUNDERSTORMW WINDS", "THUNDERSTORMWINDS", "THUNDERSTROM WIND", "THUNDERSTROM WINDS",
"THUNDERSTORM WINDS", "THUNDERTORM WINDS", "THUNDERTSORM WIND", "THUNDESTORM WINDS", and "THUNERSTORM WINDS".
In this analysis we make no effort to combine these.